### 2017-05-04 2017mergingmerges/section/XX-Related.tex
initial benchmarks
serious cutting

@@

Recent work on stream fusion by \citet{kiselyov2016stream} uses staged computation in a push-based system to ensure all combinators are inlined, but when streams are used multiple times this causes excessive inlining, which duplicates work.
For effectful inputs such as reading from the network, duplicating work changes the semantics.
% I could write more about this eg only supporting a single output, but the other points probably apply to push streams in general

@@

One way to address the difference between pull and push streams is to explicitly support both separately, as seen in \citet{bernardy2015duality} and \citet{lippmeier2016polarized}.
Here, pull streams have the type @Source@ and represent a source that is always available to be pulled from, while push streams have the type @Sink@ and represent a sink that can always accept input.
Both systems rely on stream bindings being used linearly to ensure correctness, including boundedness of buffers.
Operations over sources are expressed fairly naturally compared to streams, for example the @zip@ combinator has the type @Source a -> Source b -> Source (a,b)@.
Sinks, however, are co-variant, and operations must be performed somewhat backwards, so that the @unzip@ combinator takes the two output sinks to push into and returns a new sink that pushes into these.
It has the type @Sink a -> Sink b -> Sink (a,b)@.
This system requires the streaming computation to be manually split into sources and sinks, and be joined together by a loop that `drains' values from the source and pushes them into the sink.

The duality between pull and push arrays has also been explored in Obsidian by \citep{claessen2012expressive} and later in \citep{svensson2014defunctionalizing}.
Here the distinction is made for the purpose of code generation for GPUs rather than fusion, as operations such as appending pull arrays require conditionals inside the loop, whereas using push arrays moves these conditionals outside the loop.

@@

In relation to process calculi, synchronised product has been suggested as a method for fusing Kahn process networks together~\cite{fradet2004network}, but does not appear to have been implemented or evaluated.
The synchronised product of two processes allows either process to take independent or local steps at any time, but shared actions, such as when both processes communicate on the same channel, must be taken in both processes at the same time.
This is a much simpler fusion method than ours, but is also much stricter.
When two processes share multiple channels, synchronised product will fail unless both processes read the channels in exactly the same order.
Our system can be seen as an extension of synchronised product that allows some leeway in when processes must take shared steps: they do not have to take shared steps at the same time, but if one process lags behind the other, it must catch up before the other one gets too far ahead.

@@

Synchronous dataflow (not to be confused with synchronous languages above) is a dataflow graph model of computation where each dataflow actor has constant, statically known input and output rates.
The main advantage of synchronous dataflow is that it is simple enough for static scheduling to be decidable, but this comes at a cost of expressivity.
StreamIt~\cite{thies2002streamit} uses synchronous dataflow for scheduling when possible, otherwise falling back to dynamic scheduling~\cite{soule2013dynamic}.
Boolean dataflow and integer dataflow~\cite{buck1993scheduling,buck1994static} extend synchronous dataflow with boolean and integer valued control ports, and attempt to recover the structure of ifs and loops from select and switch actors.
These systems allow some dynamic structures to be scheduled statically, but are very rigid and only support limited control flow structures: it is unclear how merge or append could be scheduled by this system.
Finite state machine-based scenario aware dataflow (FSM-SADF)~\cite{stuijk2011scenario,van2015scenario} is still quite expressive compared to boolean and integer dataflow, while still ensuring static scheduling.
A finite state machine is constructed, where each node of the FSM denotes its own synchronous dataflow graph.
The FSM transitions from one dataflow graph to another based on control outputs of the currently executing dataflow graph.
For example, a filter is represented with two nodes in the FSM.
The dataflow graph for the initial state executes the predicate, and the value of the predicate is used to determine which transition the FSM takes: either the predicate is false and the FSM stays where it is, or the predicate is true and moves to the next state.
The dataflow graph for the next state emits the value, and moves back to the first state.
This does appear to be able to express value-dependent operations such as merge, but lacks the composability - and familiarity - of combinators.

>>>>

\TODO{Cite meta-repa \cite{ankner2013edsl}}

@@

Recent work on stream fusion \cite{kiselyov2016stream} uses staged computation in a push-based system to ensure all combinators are inlined, but when streams are used multiple times this causes excessive inlining, which duplicates work.
For effectful inputs such as reading from the network, duplicating work changes the semantics.
% I could write more about this eg only supporting a single output, but the other points probably apply to push streams in general

@@

One way to address the difference between pull and push streams is to explicitly support both \cite{bernardy2015duality, lippmeier2016polarized}.
Here, pull streams have the type @Source@ and represent a source, always ready to be pulled, while push streams have the type @Sink@ and represent a sink, always ready to accept data.
% Both systems rely on stream bindings being used linearly to ensure correctness.
Operations over sources are expressed fairly naturally compared to streams, for example the @zip@ combinator has the type @Source a -> Source b -> Source (a,b)@.
Sinks are co-variant, and operations must be performed somewhat backwards, so that the @unzip@ combinator takes the two output sinks to push into and returns a new sink that pushes into these.
% It has the type @Sink a -> Sink b -> Sink (a,b)@.
This system requires the streaming computation to be manually split into sources and sinks.
% and be joined together by a loop that `drains' values from the source and pushes them into the sink.

The duality between pull and push arrays has also been explored in Obsidian \cite{claessen2012expressive, svensson2014defunctionalizing}.
Here the distinction is made for the purpose of code generation for GPUs rather than fusion, as operations such as appending pull arrays require conditionals inside the loop, whereas using push arrays moves these conditionals outside the loop.

@@

In relation to process calculi, synchronised product has been suggested as a method for fusing Kahn process networks together~\cite{fradet2004network}, but does not appear to have been implemented or evaluated.
The synchronised product of two processes allows either process to take independent or local steps at any time, but shared actions, such as when both processes communicate on the same channel, must be taken in both processes at the same time.
This is a much simpler fusion method than ours, but is also much stricter.
When two processes share multiple channels, synchronised product will fail unless both processes read the channels in exactly the same order.
Our system is an extension of synchronised product that allows some leeway in when processes must take shared steps: they do not have to take shared steps at the same time, but if one process lags behind the other, it must catch up before the other one gets too far ahead.

@@

Synchronous dataflow (SDF; not to be confused with synchronous languages above) is a dataflow graph model of computation where each node has constant, statically known input and output rates.
SDF is simple enough for static scheduling to be decidable, but this comes at a cost of expressivity.
StreamIt~\cite{thies2002streamit} uses synchronous dataflow for scheduling when possible, otherwise falling back to dynamic scheduling~\cite{soule2013dynamic}.
Boolean dataflow and integer dataflow~\cite{buck1993scheduling,buck1994static} extend SDF with boolean and integer valued control ports, but only support limited control flow structures.
Finite state machine-based scenario aware dataflow (FSM-SADF)~\cite{stuijk2011scenario,van2015scenario} is quite expressive compared to boolean and integer dataflow, while still ensuring static scheduling.
A finite state machine is constructed, where each node of the FSM denotes its own SDF graph.
The FSM transitions from one dataflow graph to another based on control outputs of the currently executing dataflow graph.
For example, a filter is represented with two nodes in the FSM.
The initial state executes the predicate, and the value of the predicate is used to determine which transition the FSM takes: either the predicate is false and the FSM stays where it is, or the predicate is true and moves to the next state.
The next state emits the value, and returns to the start.
This appears to be able to express value-dependent operations such as merge, but lacks the composability and familiarity of combinators.
